---
title: "Factors Influencing the Length of Stay for Shelter Animals"
author: "Group 18"
format: 
  html:
    embed-resources: true
  pdf: default
editor_options: 
  chunk_output_type: inline
execute:
  echo: false
  eval: true
  warning: false
  message: false
  geometry: margin=1in,landscape
header-includes:
  - \usepackage{float}  
  - \floatplacement{figure}{H}  
  - \floatplacement{table}{H}
---

Importing necessary libraries and loading the dataset:

```{r}
library(ggplot2)
library(dplyr)
library(moderndive)
library(gapminder)
library(skimr)
library(MASS)
library(AER)
library(pscl)
animal_data<- read.csv("dataset18.csv")
```

## Introduction

The dataset, acquired from the Dallas animal shelter, encompasses comprehensive information regarding each animal admitted to the facility. It contains vital details such as the type of animal namely {dogs, cats, wildlife, bird, livestock}, the reason for admission, and the eventual outcome i.e whether they were euthanised, returned to the owner, adopted, foster or died. This study aims to investigate a fundamental question:

-   **What are the factors that influence the time an animal spends at shelter before their final outcome**

## Exploratory Data Analysis

Taking a closer look at the dataset:

```{r}
head(animal_data)
```

```{r}
glimpse(animal_data)
```

Let's look at the summary statistic for our response variable i.e the time an animal spends at a shelter before their final outcome.

```{r}
animal_data[,'time_at_shelter'] %>%
  skim()
```

The mean time spent is around six days. Additionally, 25% of the animals stayed less for one day and less. Also, 75% of the animals present stayed for nine days and less

The distribution(hist plot) is right skewed depicting most animals spent quite less time in general.

```{r}
ggplot(animal_data, mapping = aes(x=time_at_shelter))+
  geom_histogram(binwidth = 5, color="black")+
  labs(title = "Distribution of response variable", x="Time at shelter", y="Count")
```

Now, let's look at the correlation between time of the year and the time an animal spent at the shelter.

```{r}
animal_data %>%
  dplyr::select(month, time_at_shelter)%>%
  get_correlation(formula = time_at_shelter~month)
```

The above coefficient states that there is very weak positive correlation between an animals time at shelter and the month of an year.

Looking at the summary stats for response when grouped by the type of animals-

```{r}
animal_data%>%
  group_by(animal_type) %>%
  summarize(count=n(), median = median(time_at_shelter), mean = mean(time_at_shelter), Std_Dev = sd(time_at_shelter),quantile_25 = quantile(time_at_shelter, 0.25), quantile_75 = quantile(time_at_shelter, 0.75))
```

Boxplot depicting the summary stats above can be seen as-

```{r}
ggplot(animal_data, mapping = aes(x=animal_type, y=time_at_shelter, fill = animal_type)) +
  geom_boxplot()+
  labs(x = "Type of Animal", y = "Time Spent at Shelter", title = "Time spent at shelter by different types of animal") +
  scale_fill_viridis_d(option = "G") 
```

It can be clearly seen that both BIRD and LIVESTOCK spent zero days at the shelter, whereas there are some significant outliers for DOGS and CATS where median time is four and three days respectively but some recorded dogs and cats spent more than twenty days and the extremes being around seventy days.

```{r}
ggplot(data=animal_data, aes(x=intake_type, y=time_at_shelter, fill=outcome_type))+
  geom_col(position = "dodge")+
  labs(x = "Intake type by outcome", y = "Time Spent at Shelter", title = "Time spent at shelter vs Outcome of the animals & their intake") +
  scale_fill_viridis_d(option = "G")
```

It can be seen from the column graph that the animal that spent the most time at shelter where the intake was owner surrendering was Euthanize and where intake was stray they were adopted at the end.

## Remove irrelevant categories

```{r}
#Removing irrelavant/outnumbered categories

animal.sub<- animal_data%>%
  filter(animal_type!='BIRD' & animal_type !='LIVESTOCK')
```

```{r}
ggplot(animal.sub, mapping = aes(x=animal_type, y=time_at_shelter, fill = animal_type)) +
  geom_boxplot()+
  labs(x = "Type of Animal", y = "Time Spent at Shelter", title = "Time spent at shelter by selected animal types") +
  scale_fill_viridis_d(option = "G") 
```

## Model fit

From the explanatory analysis part we can see that our response variable is count data and the distribution shows an exponential trend. Therefore Poisson regression and negative binomial regression are potentially suitable for the data.

```{r}
#Poisson regression model fitting
model_poi<-glm(time_at_shelter~animal_type+intake_type+chip_status, family = poisson(link = "log"), data=animal.sub)
summary(model_poi)
```

From the model result, we conclude that only the coefficient of animal_type at level "WILDLIFE" is not statistically significant under the 95% confidence level, meaning when animal type changes from cat to wildlife the time at shelter is not expected to have a significant change.

We carry out a chi-square test to see if the model is a good fit and return the p-value.

```{r}
#Chi-square test for the Poisson model
chi_poi <- pchisq(model_poi$null.deviance-model_poi$deviance, df=model_poi$df.null-model_poi$df.residual, lower.tail = FALSE)
print(chi_poi)
```

The p-value in the chi-square test is lower than 0.05, suggesting the model is a good fit.

```{r}
#Residual plot of the poisson model
residual_poi<-data.frame(fitted_poi=model_poi$fitted.values,resid_poi=residuals(model_poi, type = "deviance"))
ggplot(residual_poi,aes(x = fitted_poi, y=resid_poi)) +
  geom_point(color = "steelblue",alpha=0.5)+
  geom_hline(aes(yintercept = mean(resid_poi)), color = "red", linetype = "dashed")+
  labs(title = "Residuals of Poisson Model",
       x="Fitted Value",
       y="Residual")
```

From the plot, we can see the residual deviances of the Poisson model seem to have 0 mean. But the variance seems larger on the positive side.

However, Poisson regression assumes that the mean and variance of the response variable are equal. We use a dispersion test to test if the Poisson regression is appropriate for our dataset.

```{r}
#Dispersion test
dispersion_test <- dispersiontest(model_poi)
print(dispersion_test)
```

As we can see the p-value is smaller than 0.05, suggesting it is not appropriate to use a Poisson model due to the overdispersion of the response variable.

```{r}
#overdispersion check
ggplot(model_poi, aes(x=log(fitted(model_poi)), y=log((animal.sub$time_at_shelter-fitted(model_poi))**2)))+ 
  geom_point(col="steelblue") +
  geom_abline(slope=1, intercept=0, col="red", size=0.5, lty = "dashed") + 
  ylab(expression((y-hat(mu))**2)) + xlab(expression(hat(mu)))
```

If Var(𝑌𝑖) \> 𝜇𝑖 we say that we have **overdispersion**. This appears to be the case for the animals data – notice in the figure above that lot of the points lie above the line of equality for mean and variance.

Therefore we try to fit a negative binomial model to overcome this problem.

```{r}
#Negative binomial model fitting
model_nb1<-glm.nb(time_at_shelter~animal_type+intake_type+chip_status,data=animal.sub)
summary(model_nb1)
```

```{r}
#F test to determine the significance of the regression coefficients
X2 <- sum(resid(model_nb1, type = "deviance")**2) 
dp <- X2 / model_nb1$df.res
dp
summary(model_nb1, dispersion = dp)
```

With the use of the estimated dispersion parameter the Wald tests are not very reliable, so we turn to an F test to determine the significance of the regression coefficients:

```{r}
drop1(model_nb1, test = "F")
```

Judging from the p-value, all levels of animal type are not significant at 95% confidence level. We take out animal type and fit a new negative binomial model.

```{r}
#Negative binomial model fitting without the insignificant variable
model_nb2<-glm.nb(time_at_shelter~intake_type+chip_status,data=animal.sub)
summary(model_nb2)
```

From the summary, we can see that only the coefficient of chip_status at level "SCAN NO CHIP" is not statistically significant, meaning when chip status changes from scan chip to scan no chip the time at shelter is not expected to have a significant change.

We carry out a chi-square test to see if the model is a good fit and return the p-value.

```{r}
#Chisqure test for the negative binomial model
chi_nb <- pchisq(model_nb2$null.deviance-model_nb2$deviance, df=model_nb2$df.null-model_nb2$df.residual, lower.tail = FALSE)
print(chi_nb)
```

The p-value in the chi-square test is lower than 0.05, suggesting the model is a good fit.

```{r}
#Residual plot of the negative binomial model
residual_nb<-data.frame(fitted_nb=model_nb2$fitted.values,resid_nb=residuals(model_nb2, type = "deviance"))
ggplot(residual_nb,aes(x = fitted_nb, y=resid_nb)) +
  geom_point(color = "steelblue",alpha=0.5)+
  geom_hline(aes(yintercept = mean(resid_nb)), color = "red", linetype = "dashed")+
  labs(title = "Residuals of Negative Binomial Model",
       x="Fitted Value",
       y="Residual")
```

As shown in the plot, it seems fair to say that the residuals have 0 mean and constant variance.

By comparing the Poisson model and the second negative binomial model, we can see that the AIC value for the second negative binomial model is lower, and the residuals are smaller and more constant.

Therefore we consider our second negative binomial model as our final model.

The intercept of the negative binomial model indicates that when intake type is confiscated, outcome type is adoption, chip status is scan chip, log(time at shelter) is expected to be 2.5, corresponding to the time at shelter of 12.17 days on average.

Holding all other variables in the model constant, the coefficient of intake_type "OWNER SURRENDER" indicates that when the intake type is surrender, log(time at shelter) is expected to decrease by -0.79 compared to when intake type is confiscated, corresponding to a 54.54% decrease in time at shelter on average.

Holding all other variables in the model constant, the coefficient of intake_type "STRAY" indicates that when the intake type is stray, log(time at shelter) is expected to decrease by -0.52 compared to when intake type is confiscated, corresponding to a 40.36% decrease in time at shelter on average.

Judging from the p-value, the coefficient of chip_status "SCAN NO CHIP" is not statistically significant under the 95% confidence level, which means that holding all other variables in the model constant, the time at shelter when animal chip status is scan no chip is not expected to be significantly different than when animal chip status is scan chip.

Holding all other variables in the model constant, the coefficient of chip_status "UNABLE TO SCAN" indicates that when the chip status is unable to scan, log(time at shelter) is expected to decrease by -0.71 compared to when ship status is scan chip, corresponding to a 50.97% decrease in time at shelter on average.

### Model from Advanced Predictive Modelling

Sometimes overdispersion is observed due to excess zeros in the data. It can be seen that our model has about 25% of observation consisting of zero values which is a significant amount. The zeros represent **immediate outcomes**, which are structurally different from positive shelter times.

### Zero-Inflated model

In zero-inflated negative binomial models, we assume that there are two processes. One is a logistic process(immediate return or not) and the other models the non-zero counts with a neg binomial process and the resulting data distribution is a mixture of the two.

```{r}
zinb_model <- zeroinfl(time_at_shelter ~ animal_type + intake_type + chip_status, dist = "negbin", data = animal.sub)
summary(zinb_model)
```

The model comparison by AIC value is carried out.

```{r}
#Model Comparison
AIC(model_nb2, zinb_model)
```

It can be seen that the model now fits the data slightly better when comparing the AIC values.
